Extending the WebID Protocol with Access Delegation

### 2012-09-07 paper.tex
simplifications in the text. (should have been part of previous comit)

\begin{abstract}
The WebID protocol enables the global identification and authentication of agents in a distributed manner by combining asymmetric cryptography and Linked Data.
In order to decide whether access should be granted or denied to a particular WebID, the authenticating web server may need to retrieve other profiles and linked resources to work out e.g. if the requesting agent is member of an authorized group.
If these resources are required to be publicly available for the server to access it, then this would be a major privacy limitation on a linked Social Network.
In this paper we explore different ways in which an agent can act as a user and we propose an extension to the WebID protocol which allows for delegation of access authorization from a WebID to a third party, e.g. allowing a server to be able to act on behalf of its users.
This extends the range of application scenarios where WebID authentication can be efficiently deployed while increasing privacy.
\end{abstract}

@@

The World Wide Web is a peer to peer communication system designed from the outset to work in a global space, between agents distributed around the world and acting in parallel, with no center of control, in a space where new agents can at any point join and there is no complete overview of the whole system.
These agents need know nothing of each other up to the point of their interaction.
This is the force that leads the web to its declarative functional architecture, with its emphasis on naming and a logic of unalterability of the meaning of names (URIs).

@@

The WebID protocol enables the global identification of agents using asymmetric cryptography in a way that fits cleanly with this architecture:
namely in such a way that agents can verify each others identity without having had any previous interactions and in such a way as to allow trust to build up in a decentralized manner.

A \textit{WebID} is a URI that refers to an agent - person, robot, group or other thing that can have intentions.
The WebID should be a URI which when dereferenced returns a representation whose description uniquely identifies the agent as the controller of a public key~\cite{sporny-m-2011--a,story-h-2009--a}

The WebID protocol is used currently mostly for client authentication%
\footnote{Server authentication using IETF DANE follows much the same logic, except that the lookup for the identity is not done using the HTTP protocol but DNSSEC~\cite{hoffman-p-2012--a}.}.
It is worth noting that the host serving the WebID profiles controls the identity of every agent whose URI is within that server's namespace.
This service is known as the origin server~\cite{barth-a-2011--a}.
It is the origin of all resources served by it.

We can easily think of the origin server as not only able to respond to requests, but also as an agent able to make requests.
Indeed WebID authentication requires the server to make WebID profile requests to other servers in order to verify the identity of agents making a request to it.
The WebID specification describes this task as being accomplished by a separate agent, the WebID verifier - which could indeed be done by another service on the web (WebID proxy authentication servers).
But it can also be so closely tied to the web service that it would be natural to think of it as part of that same service.
The WebID profile furthermore could be served by the same agent as the one making the request, in which case we have a minimal case of a peer to peer communication.
Note that fetching a WebID profile for WebID authentication should be done anonymously, for fear of authentication deadlocks%
\footnote{For example one can imagine an agent $S$ with profile $P_s$ requesting a resource on server $R$ which requires authentication.
$S$ would send $R$ its certificate, thereby requiring $R$ to dereference $S$'s profile $P_s$ in order to verify the WebID.
If $P_s$ itself requires authentication of $R$ and if $R$ sends a certificate containing a WebID with its profile $P_r$, and if $P_r$ itself requires authentication then it looks like we have a deadlock.}.

Things get more interesting in the authorization space.
Consider a very natural application of WebID: allowing friends of one's friends access to some resources.
This authorization rule will require the web server to fetch each of its users' friends profiles, in order to build up the list of authorised users.
But there is a privacy issue involved here: not everyone wants to make all of their social network publicly visible, and some may not want to make any of it publicly visible.
Those people may then protect their FOAF profile with access control rules such as only allowing friends of their friends access to it.
How can a server that needs access to these FOAF profiles in order to apply its own access control rules get access to the information? 
Would the server itself need to be listed as a friend of a friend by each of the users friends?
Should the server take on the identity of the user it is fetching resources for? 
How would it be able to do so?
What other solutions are there?
These are the questions we will try to answer in this paper.

@@

When the user owns the origin server and there is a one-to-one relation between the two, as it is the case in a personal FreedomBox\footnote{http://www.freedomboxfoundation.org/}, then the origin server is just an extension of the user, and in that case trusting the server is the same as trusting the user.
%In that case the origin server can in fact just authenticate as the user whenever needed.

@@

Still, such an agent connecting to a remote service on behalf of one of its users would then need to identify which user it was acting on behalf of, or the remote service receiving a request would not know which access control rule to apply to the requesting origin server.
For example in our previous example in the Romeo and Juliet story we showed how the Capulet family FreedomBox only shows Juliet as knowing Romeo, when Romeo makes the request on the resource, but not when the request is made by Romeo's father, lord Montague.
That was an easy access control rule for the Capulet FreedomBox to make when the request was using the WebID protocol, since there it would know via the WebID authentication process exactly which identity the user was making the request as. 
But if it is the Origin Server connecting to the Capulet (FreedomBox), identified as their Montague Origin Server, then what representation should the Capulet FreedomBox return? 
Should it return the reduced version available to Lord Montague?
Or should it return the more complete version available to Romeo and his close friends?

The proposal here is for the Origin Server to add to each HTTP request (made over TLS) an \lstinline|On-Behalf-Of| header identifying the user on behalf of which the request is being made with that user's WebID, as shown below:

\begin{verbatim}
GET /juliet/knows HTTP/1.1
Host: capulet.org
On-Behalf-Of: https://montague.net/romeo#me
User-Agent: FreedomBox_FamilyEdition/0.1 
Accept: application/rdf+xml,text/turtle,application/xhtml+xml
\end{verbatim}

The Capulet FreedomBox Guard having verified that this request does indeed come from the origin server of Romeo's profile
\lstinline|<https://montague.net/romeo>| would then be able to accept that since the requesting server could in any case act as Romeo whenever it chose to do so, this protocol making exchange between the Montague and Capulet houses more efficient without making it less secure, this be justification enough for it (the Capulet FreedomBox) to  serve the representation that Romeo would have received had he connected to the server directly.  
In \autoref{sec:relatedwork} we will argue that this is very similar to the way CORS deals with access control, namely by adding a header to a request in order to clarify what agent is making the request.

This \lstinline|On-Behalf-Of| header makes it possible for the \textit{secretary} to open only one TLS connection to a remote server and be 
able to specify for each request on behalf of whom it is making it. The Montague FreedomBox could make a request to the Capulet Freedom Box \lstinline|On-Behalf-Of| Romeo, a few milliseconds later \lstinline|On-Behalf-Of| Lord Montague, and so on. 
Again as explained in the previous section it is the responsibility of the Montague FreedomBox to keep the returned information separate, and act with discretion whenever it is acting on behalf of one or the other of the users. 
As such it's role is not unlike the role of Friar Laurence in the play - dedication to people's private matters requires tact and discretion.

@@

If we now identify the secretary that can act on behalf of a user using a WebID, we can generalise the protocol somewhat.
That is, when an agent - such as the Montague FreedomBox Laurence with WebID \lstinline|<http://montague.net/fb#laurence>| - authenticates to a remote server it can use its own WebID. 
This would allow the \textit{secretary} to have her own public key, and so to reduce the risk of her private key being compromised affecting the server public key.

But how would the remote server know that it can trust that secretary to be acting on behalf of a particular user?
It can no longer just compare the TLS keys of the requesting agent to see if it comes from the same Origin Server.
We need to make this relation explicit by use of a special RDF relation provisionally called \lstinline|:secretary|%
\footnote{A object property with a domain and range as \lstinline|foaf:Agent| which we will provisionally place
in the \lstinline|cert:| namespace, though it may be more appropriately placed in the \lstinline|auth:| namespace.}

@@

The remote server can then verify that the identified agent is the secretary of the agent he wishes to act on behalf of (as specified in the \lstinline|On-Behalf-Of| header, by dereferencing that user's profile and verifying that the user specifies the \texttt{:secretary} relation there, as it would if Romeo had the FOAF profiles as shown in \autoref{listing:webid1}.

@@

%\autoref{fig:AuthSequence} depicts the process of a WebID authentication which is extended with the following components to allow access delegation:
The following enumeration describes each authentication step of \autoref{fig:AuthSequence} in detail but concentrates on the context of access delegation:
(1) The secretary  opens a TLS connection with the server of the protected resource.
(2) Once TLS is set up, the HTTP request is sent to the server (e.g. a \verb!HTTP GET!), with an additional \lstinline|On-Behalf-Of| header, which thereby defines the requesting agent as a \textit{secretary} and the referred to agent as the \textit{principal} 
(3) The guard intercepts this request, and  in turn requests client authentication using TLS session renegotiation.
The \textit{secretary} authenticates as itself by sending a Certificate containing a WebID referring to it.
The TLS-Light service verifies that the \textit{secretary} really is in possession of the private key corresponding to the public key sent in the certificate.
This is defined in the TLS protocol \cite{dierks-t-2012--a}.
(4) The guard ask the verification agent to verify the secretary WebID which is named in the certificate (5a).
This process is exactly as described in the WebID protocol~\cite{story-h-2009--a}.
The guard also asks the verifier agent to check the secretary claim implied by the \lstinline|On-Behalf-Of| header.
(5b) The \textit{principal} agent's relation to the \textit{secretary} is  verified by dereferencing the \textit{principal}'s WebID Profile, and verifying it responds with a true to the SPARQL ASK query \lstinline|ASK { ?principal :secretary ?secretary . }| where the \lstinline|?principal| and \lstinline|?secretary| variables have been bound to the correct URIs.
(6) The authentication and verification process having succeeded, the authorization process checking if the \textit{principal} would get access to the requested resource.
(7) The resource representation can then be returned or not depending on the access control rules.

@@

\paragraph{Cross Origin Resource Sharing (CORS)}
In this paper we proposed to add an \lstinline|On-Behalf-Of| header to each HTTP request made by a \textit{secretary} in order to be able to identify the \textit{principal} she is working for in that particular request.
This is quite a major semantic addition to HTTP, but for which there is a widely adopted precedent, namely CORS\footnote{\url{http://www.w3.org/TR/cors/}}, the Cross Origin Resource Sharing specification at the W3C.
CORS identifies the remote \textit{principal} with an \lstinline|Origin| header, and uses any HTTP authentication method to identify the \textit{secretary} which is the browser.

@@

The use cases for CORS are slightly different from the ones for our delegation proposal. 
In CORS the browser is using its own identity credentials when connecting to remote servers (e.g. a bank), as it needs to alert the remote service that the request originates from a particular agent.
The browser also needs to make sure the remote server understands this, especially when authentication is required, to allow the remote server to evaluate the strength of the JavaScript trust relation.
In the \autoref{sec:spec} use cases, the remote site trusts the \textit{principal}, but may not be aware of the \textit{secretary}.
In our proposal, the identification of the agent in the \lstinline|On-Behalf-Of| header is much more precise.
Combined with the ability of the secretary to also identify herself with a WebID, it builds an explicit trust relation between the \textit{principal} and the \textit{secretary}. The vagueness of the \lstinline|Origin| identity of the \textit{principal} in CORS makes a lot of this more difficult.

It is a topic of further research to evaluate in which circumstances it is important for the \lstinline|secretary| to also make sure that the remote server is aware that the connection being made is \lstinline|On-Behalf-Of| the specified \textit{principal}. 
For otherwise it is possible that the \textit{secretary} herself be made liable for the actions she is making on behalf of her \textit{principal}, especially were she to make \texttt{PUT}, \texttt{POST} or \texttt{DELETE} requests. 
The CORS protocol determines this by  requiring an initial \texttt{HEAD} request to a resource to be made, which will determine whether the server understands CORS.
With the help of TLS the server hosting the remote resource could have a WebID Profile describing it to be an understander of the \lstinline|On-Behalf-Of| header, making the protocol a lot more efficient and secure once more.

@@

%\todo{Nice up wording esp. our terms and language}
In this paper we discussed different ways in which an agent, such as a Social Web application, can act on behalf of a user and request as well as manipulate resources on the Web in the name this user.
Our proposal for supporting such a communication schema involves an extension to the WebID protocol, which allows for delegation of access authorization from one agent to another agent, both identified by their WebID.
The technical solution for this concept annotates HTTP requests by adding the \lstinline|On-Behalf-Of| header field which refers to the WebID of the user in whose name the agent acts.
This distinguish the identities of the user and the agent and allows for creation of policy descriptions which use both WebIDs to specify the relations between the agent and the user in order to setup the access rules for this agent.
Since the protocol footprint is minimal, the main information entropy to decide whether an agent has the right to request a specific resource or not is available as part of the content of a WebID profile which can be requested as Linked Data.
We believe that such an shift of the entropy from a protocol endpoint (such as in OAuth) to a machine readable document is a crucial feature of Linked Data in general and WebID in detail.

>>>>

\begin{abstract}
The WebID protocol enables the global identification and authentication of agents in a distributed manner by combining asymmetric cryptography and Linked Data.
In order to decide whether access should be granted or denied to a particular WebID, the authenticating web server may need to retrieve other profiles and linked resources to work out if the requesting agent is member of an authorized group (e.g. friends of the resource owner's friends).
If it were required for such resources to be publicly available, this would be a major privacy limitation on a linked Social Network.
In this paper we explore different ways in which an agent can act on behalf of a user and we propose an extension to the WebID protocol which allows for delegation of access authorization.
This extends the range of application scenarios where WebID authentication can be efficiently deployed while increasing privacy.
\end{abstract}

@@

The World Wide Web is a peer to peer communication system designed from the outset to work in a global space, between agents distributed around the world and acting independently, with no center of control, in a space where new agents can at any point join and there is no complete overview of the whole system.
These agents need know nothing of each other up to the point of their interaction.
These architectural requirements are the reason for the web's declarative functional architecture, with its emphasis on naming and a logic of unalterability of the meaning of names (URIs).

@@

A \textit{WebID} is a URI that refers to an agent - person, robot, group or other thing that can act intentionally.
The WebID should be a URI which when dereferenced returns a representation whose description uniquely identifies the agent as the controller of a public key~\cite{sporny-m-2011--a,story-h-2009--a}.

The WebID protocol enables the global identification of agents using asymmetric cryptography in a way that fits cleanly with web architecture:
namely in such a way that agents can verify each others identity without having had any previous interactions and in such a way as to allow trust to build up in a decentralized manner through a linked web of trust.

It is worth noting that the host serving the WebID profiles controls the identity of every agent whose URI is within that server's namespace.
This service is known as the origin server~\cite{barth-a-2011--a}.

The WebID protocol is designed for client authentication%
\footnote{Server authentication using IETF DANE follows much the same logic, except that the lookup for the identity is not done using the HTTP protocol but DNSSEC~\cite{hoffman-p-2012--a}.}.
But the origin server does not just respond to requests, it also is able to make requests.
Indeed WebID authentication requires the server to make WebID profile requests to other servers in order to verify the identity of agents%
\footnote{Note that fetching a WebID profile for WebID authentication should be done anonymously, for fear of authentication deadlocks. 
For example one can imagine an agent $S$ with profile $P_s$ requesting a resource on server $R$ which requires authentication.
$S$ would send $R$ its certificate, thereby requiring $R$ to dereference $S$'s profile $P_s$ in order to verify the WebID.
If $P_s$ itself requires authentication of $R$ and if $R$ sends a certificate containing a WebID with its profile $P_r$, and if $P_r$ itself requires authentication then it looks like we have a deadlock.}.

Things get more interesting in the authorization space.
Consider a very natural application of WebID: allowing friends of one's friends access to some resources.
This authorization rule will require the web server to fetch each of the resource owner's friends profiles, in order to build up the list of authorised users.
But there is a privacy issue involved here: not everyone wants to make all of their social network publicly visible, and some may not want to make any of it publicly visible.
Those people may then protect their FOAF profile with access control rules such as one only allowing friends of their friends access to it.
How can a server that needs access to these FOAF profiles in order to apply its own access control rules get access to the information? 
Would the server itself need to be listed as a friend of a friend by each of the users friends?
Should the server take on the identity of the user it is fetching resources for? 
How would it be able to do so?
What other solutions are there?
These are the questions we will try to answer in this paper.

@@

When the user owns the origin server and there is a one-to-one relation between the two, as it is the case in a personal FBx\footnote{\url{http://www.freedomboxfoundation.org/}} (FBx henceforth), then the origin server is just an extension of the user, and in that case trusting the server is the same as trusting the user.
%In that case the origin server can in fact just authenticate as the user whenever needed.

@@

Still, such an agent connecting to a remote service on behalf of one of its users would then need to identify which user it was acting on behalf of, or the remote service receiving a request would not know which access control rule to apply to the requesting origin server.
For example in our previous example in the Romeo and Juliet story we showed how the Capulet family FBx only shows Juliet as knowing Romeo, when Romeo makes the request on the resource, but not when the request is made by Romeo's father, lord Montague.
That was an easy access control rule for the Capulet FBx to make when the request was using the WebID protocol, since there it would know via the WebID authentication process exactly which identity the user was making the request as. 
But if it is the Origin Server connecting to the Capulet FB, identified as their Montague Origin Server, then what representation should the Capulet FBx return? 
Should it return the reduced version available to Lord Montague?
Or should it return the more complete version available to Romeo and his close friends?

The proposal here is for the Origin Server to add to each HTTP request (made over TLS) an \lstinline|X-On-Behalf-Of| header identifying the user on behalf of which the request is being made with that user's WebID, as shown below:

\begin{verbatim}
GET /juliet/knows HTTP/1.1
Host: capulet.org
X-On-Behalf-Of: https://montague.net/romeo#me
User-Agent: FreedomBox_FamilyEdition/0.1 
Accept: application/rdf+xml,text/turtle,application/xhtml+xml
\end{verbatim}

The Capulet FBx Guard having verified that this request does indeed come from the origin server of Romeo's profile
\lstinline|<https://montague.net/romeo>| would then be able to accept that since the requesting server could in any case act as Romeo whenever it chose to do so, this protocol making exchange between the Montague and Capulet houses more efficient without making it less secure, this be justification enough for it (the Capulet FBx) to  serve the representation that Romeo would have received had he connected to the server directly.  
In \autoref{sec:relatedwork} we will argue that this is very similar to the way CORS deals with access control, namely by adding a header to a request in order to clarify what agent is making the request.

This \lstinline|X-On-Behalf-Of| header makes it possible for the \textit{secretary} to open only one TLS connection to a remote server and be 
able to specify for each request on behalf of whom it is making it. The Montague FBx could make a request to the Capulet Freedom Box \lstinline|X-On-Behalf-Of| Romeo, a few milliseconds later \lstinline|X-On-Behalf-Of| Lord Montague, and so on. 
Again as explained in the previous section it is the responsibility of the Montague FBx to keep the returned information separate, and act with discretion whenever it is acting on behalf of one or the other of the users. 
As such it's role is not unlike the role of Friar Laurence in the play - dedication to people's private matters requires tact and discretion.

@@

If we now identify the secretary that can act on behalf of a user using a WebID, we can generalise the protocol somewhat.
That is, when an agent - such as the Montague FBx Laurence with WebID \lstinline|<http://montague.net/fb#laurence>| - authenticates to a remote server it can use its own WebID. 
This would allow the \textit{secretary} to have her own public key, and so to reduce the risk of her private key being compromised affecting the server public key.

But how would the remote server know that it can trust that secretary to be acting on behalf of a particular user?
It can no longer just compare the TLS keys of the requesting agent to see if it comes from the same Origin Server.
We need to make this relation explicit by use of a special RDF relation provisionally called \lstinline|:secretary|%
\footnote{\url{http://bblfish.net/work/2012/09/trust#secretary} }

@@

The remote server can then verify that the identified agent is the secretary of the agent he wishes to act on behalf of (as specified in the \lstinline|X-On-Behalf-Of| header, by dereferencing that user's profile and verifying that the user specifies the \texttt{:secretary} relation there, as it would if Romeo had the FOAF profiles as shown in \autoref{listing:webid1}.

@@

%\autoref{fig:AuthSequence} depicts the process of a WebID authentication which is extended with the following components to allow access delegation:
The following enumeration describes each authentication step of \autoref{fig:AuthSequence} in detail but concentrates on the context of access delegation:
(1) The secretary  opens a TLS connection with the server of the protected resource.
(2) Once TLS is set up, the HTTP request is sent to the server (e.g. a \verb!HTTP GET!), with an additional \lstinline|On-Behalf-Of| header, which thereby defines the requesting agent as a \textit{secretary} and the referred to agent as the \textit{principal} 
(3) The guard intercepts this request, and  in turn requests client authentication using TLS session renegotiation.
The \textit{secretary} authenticates as itself by sending a Certificate containing a WebID referring to it.
The TLS-Light service verifies that the \textit{secretary} really is in possession of the private key corresponding to the public key sent in the certificate.
This is defined in the TLS protocol \cite{dierks-t-2012--a}.
(4) The guard asks the verification agent to verify the secretary WebID which is named in the certificate (5a).
This process is exactly as described in the WebID protocol~\cite{story-h-2009--a}.
The guard also asks the verifier agent to check the secretary claim implied by the \lstinline|X-On-Behalf-Of| header.
(5b) The \textit{principal} agent's relation to the \textit{secretary} is  verified by dereferencing the \textit{principal}'s WebID Profile, and verifying it responds with a true to the SPARQL ASK query \lstinline|ASK { ?principal :secretary ?secretary . }| where the \lstinline|?principal| and \lstinline|?secretary| variables have been bound to the correct URIs.
(6) The authentication and verification process having succeeded, the authorization process checking if the \textit{principal} would get access to the requested resource.
(7) The resource representation can then be returned or not depending on the access control rules.

@@

\paragraph{Cross Origin Resource Sharing (CORS)}
In this paper we proposed to add an \lstinline|X-On-Behalf-Of| header to each HTTP request made by a \textit{secretary} in order to be able to identify the \textit{principal} she is working for in that particular request.
This is quite a major semantic addition to HTTP, but for which there is a widely adopted precedent, namely CORS\footnote{\url{http://www.w3.org/TR/cors/}}, the Cross Origin Resource Sharing specification at the W3C.
CORS identifies the remote \textit{principal} with an \lstinline|Origin| header, and uses any HTTP authentication method to identify the \textit{secretary} which is the browser.

@@

The use cases for CORS are slightly different from the ones for our delegation proposal. 
In CORS the browser is using its own identity credentials when connecting to remote servers (e.g. a bank), as it needs to alert the remote service that the request originates from a particular agent.
The browser also needs to make sure the remote server understands this, especially when authentication is required, to allow the remote server to evaluate the strength of the JavaScript trust relation.
In the \autoref{sec:spec} use cases, the remote site trusts the \textit{principal}, but may not be aware of the \textit{secretary}.
In our proposal, the identification of the agent in the \lstinline|X-On-Behalf-Of| header is much more precise.
Combined with the ability of the secretary to also identify herself with a WebID, it builds an explicit trust relation between the \textit{principal} and the \textit{secretary}. The vagueness of the \lstinline|Origin| identity of the \textit{principal} in CORS makes a lot of this more difficult.

It is a topic of further research to evaluate in which circumstances it is important for the \lstinline|secretary| to also make sure that the remote server is aware that the connection being made is \lstinline|X-On-Behalf-Of| the specified \textit{principal}. 
For otherwise it is possible that the \textit{secretary} herself be made liable for the actions she is making on behalf of her \textit{principal}, especially were she to make \texttt{PUT}, \texttt{POST} or \texttt{DELETE} requests. 
The CORS protocol determines this by  requiring an initial \texttt{HEAD} request to a resource to be made, which will determine whether the server understands CORS.
With the help of TLS the server hosting the remote resource could have a WebID Profile describing it to be an understander of the \lstinline|X-On-Behalf-Of| header, making the protocol a lot more efficient and secure once more.

@@

%\todo{Nice up wording esp. our terms and language}
In this paper we discussed different ways in which an agent, such as a Social Web application, can act on behalf of a user and request as well as manipulate resources on the Web in the name this user.
Our proposal for supporting such a communication schema involves an extension to the WebID protocol, which allows for delegation of access authorization from one agent to another agent, both identified by their WebID.
The technical solution for this concept annotates HTTP requests by adding the \lstinline|X-On-Behalf-Of| header field which refers to the WebID of the user in whose name the agent acts.
This distinguish the identities of the user and the agent and allows for creation of policy descriptions which use both WebIDs to specify the relations between the agent and the user in order to setup the access rules for this agent.
Since the protocol footprint is minimal, the main information entropy to decide whether an agent has the right to request a specific resource or not is available as part of the content of a WebID profile which can be requested as Linked Data.
We believe that such an shift of the entropy from a protocol endpoint (such as in OAuth) to a machine readable document is a crucial feature of Linked Data in general and WebID in detail.
